Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-10-25 manuscript.tex
latest

\documentclass[12pt]{article}

\usepackage{graphicx} 
\usepackage{amsmath}
%\usepackage{a4wide}
%\usepackage{multicol,caption}
%\setlength{\textwidth}{16cm}
%\setlength{\textheight}{22cm}

\begin{document}\sloppy

\title{A preliminary rcK derivation}
\author{J Bertram and J Masel}
\maketitle

@@

\section{Introduction}

@@

We broadly follow the spirit of MacArthur's r/K selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios. However, rather than building directly on MacArthur's formalism and its later extensions \cite{joshi_2001} (e.g. using Lotka-Volterra equations), our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species). This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the r/K low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities, in which case r- and K- selection will effectively coincide (more details in the Discussion).

@@

We assume that each individual in a population requires its own territory to survive and reproduce (a site-occupancy model). All territories are identical, and the total number of territories is $K$. Time $t$ advances in discrete steps, each step representing one generation. In generation $t$, the number of mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=K-N(t)$.

Each generation, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories (there is no dispersal limitation). We ignore propagules that land on occupied territories --- these are assumed to have no chance of survival and are not included in $m_i$. For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.

@@

When multiple propagules land on the same territory, they compete to secure the territory as they develop. This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next generation is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability.

The increase in $n_i$ over one generation due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories. Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $K$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$. This gives
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}

@@

\begin{figure}
\centering
\includegraphics[scale=0.7]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq. \eqref{eq:master}. $\Delta_+ n_i/m_i$ from Eq. \eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one generation over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed). Two genotypes are present. (a) and (b) show low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$).} 
\end{figure}

@@

The ideal environmental archetype is characterized by the near-absence of stress and disturbance. Consequently, $d_i\ll 1$, whereas $b_i$ can easily be of order $1$ or larger. From Eq. \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates. For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$. Moreover, since $L=b_i N/(N-K)$ by definition, population density is also high $N/K\approx 1$. Thus, territorial contests are decisively important.

The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by physical destruction. Disturbances do not only affect adults as in Eq. \eqref{eq:delttot}, but also juveniles in the process of territorial contest. These juvenile deaths can be represented as a fractional reduction in the number of territories secured. To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each generation. Then, the disturbance archetype is characterized by $d_i$ being close to $1$ (almost all adults and juveniles are killed each generation), which can only be tolerated by those organisms with large enough $b_i$ to ensure population persistence. The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, and we have $L\ll 1$ and $N/K\ll 1$. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \cite{aerts_1999}). In Grime's view, the archetypal stressful environmental imposes such severe challenges that success depends primarily on the ability to tolerate the stressors. Population density $N/K$ is suppressed to such low levels that competition between individuals is not important. In our model, this corresponds to $b_i\ll 1$ and $b_i/d_i\approx 1$. Thus $L\ll 1$ and $N/K\ll 1$ similar to the disturbance archetype, except that stress now constrains $b_i$ to be small.

The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported by the environment \cite{taylor_1990}. For example, in the case that the stress is induced by a scarcity of consumable resources, competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density. In our model, this would imply a large reduction in $K$ (greater per-individual territorial requirement). That is, $N$ under stress is much lower than  under ideal conditions, but it is not much lower than $K$ for the stressful environment. Since our model accommodates both of these alternatives, we include them both here.

@@

\begin{figure}
\centering
\begin{tabular}{*{5}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
  Parameter- & $d_i \ll 1$ & $d_i \approx 1$ & $b_i \ll 1$ & $b_i \ll 1$ \\
  regime & $b_i/d_i\gg 1$ & $b_i/d_i\gg 1$ & $b_i\approx d_i$ & $b_i>  d_i$ \\
  Density $N/K$  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$ & $\downarrow d$ & $\uparrow c$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype. Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

@@

Grime's triangle without constraints

>>>>

\documentclass[11pt]{article}
\usepackage[sc]{mathpazo}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}

\usepackage{graphicx} 
%\usepackage{amsmath}
%\usepackage{a4wide}
%\usepackage{multicol,caption}
%\setlength{\textwidth}{16cm}
%\setlength{\textheight}{22cm}

\title{Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle}
\author{Jason Bertram $^{1,\ast}$ \\ 
Joanna Masel $^{1}$}

\date{}

\begin{document}

\maketitle

\noindent{}1. Department of Ecology and Evolutionary Biology, University of Arizona, Tucson, AZ 85721.

\noindent{}$\ast$ Corresponding author; e-mail: amnat@uchicago.edu.

\bigskip

\textit{Manuscript elements}: Figure~1, figure~2, table~1, online
appendices~A and B (including figure~A1 and figure~A2). Figure~2 is to
print in color.

\bigskip

\textit{Keywords}: Examples, model, template, guidelines.

\bigskip

\textit{Manuscript type}: Article. 
% Or e-article, note, e-note, natural history miscellany,
% e-natural history miscellany, comment, reply, symposium, or
% countdown to 150.

\bigskip

\noindent{\footnotesize Prepared using the suggested \LaTeX{} 
template for \textit{Am.\ Nat.}}

\linenumbers{}
\modulolinenumbers[3]

\newpage{}

\section*{Abstract}

Some abstract stuff that runs over multiple lines so that I can see the quote formatting.

\newpage{}

@@

\section*{Introduction}

@@

We broadly follow the spirit of MacArthur's r/K selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios. However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \cite{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species). This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the r/K low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities, in which case r- and K- selection will effectively coincide (more details in the Discussion).

@@

We assume that each individual in a population requires its own territory to survive and reproduce (a site-occupancy model). All territories are identical, and the total number of territories is $K$. Time $t$ advances in discrete iterations, each representing the average time from birth to reproductive maturity. In iteration $t$, the number of reproductively mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=K-N(t)$.

Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories (no dispersal limitation). We assume adults cannot be ousted from occupied territories, so only propagules landing on occupied territories are included in $m_i$. More generally, $m_i$ does not include propagules which never even begin the development cycle. For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.

@@

When multiple propagules land on the same territory, they compete to secure the territory as they develop. This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next iteration is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability.

The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories. Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $K$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$. This gives
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}

@@

\begin{figure}
\centering
\includegraphics[scale=0.7]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq. \eqref{eq:master}. $\Delta_+ n_i/m_i$ from Eq. \eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed). Two genotypes are present. (a) and (b) show low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$).} 
\end{figure}

@@

The ideal environmental archetype is characterized by the near-absence of stress and disturbance. Consequently, $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$. From Eq. \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates. For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$. Moreover, since $L=b_i N/(N-K)$ by definition, population density is also high $N/K\approx 1$. Thus, territorial contests are decisively important.

The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by physical destruction. Disturbances do not only affect adults as in Eq. \eqref{eq:delttot}, but also juveniles in the process of territorial contest. These juvenile deaths can be represented as a fractional reduction in the number of territories secured. To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration. Then, the disturbance archetype is characterized by $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration). The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, where $b_i$ must be exceptionally large to ensure population persistence, and we have $L\ll 1$ and $N/K\ll 1$. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \cite{aerts_1999}). In Grime's view, stressful environments impose such severe challenges that surviving the stressors at all is the primary challenge. In our model, this can be expressed by $b_i\ll 1$ and $b_i/d_i\approx 1$. The propagule density $L$ (as well as $N/K$) are suppressed to such low levels that there is essentially no competition between individuals. The severity of the stress is such that mutations which appreciably improve $b_i$ are extremely rare, so $b_i$ is constrained to remain low.

The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported by the environment \cite{taylor_1990}. For example, in the commonly cited case that the stress is induced by a scarcity of consumable resources, competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density. In our model, this would imply a large reduction in $K$ (greater per-individual territorial requirement). That is, $N$ under stress is much lower than  under ideal conditions, but it is not much lower than $K$ for the stressful environment. Since our model accommodates both of these alternatives, we include them both here.

@@

\begin{figure}
\centering
\begin{tabular}{*{5}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
  Parameter- & $d_i \ll 1$ & $d_i \approx 1$ & $b_i \ll 1$ & $b_i \ll 1$ \\
  regime & $b_i/d_i\gg 1$ & $b_i/d_i\gg 1$ & $b_i\approx d_i$ & $b_i>  d_i$ \\
  Density $N/K$  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype. Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

How does Fig. 4 compare to Grime's C/S/R strategies? In disturbed environments, the ruderal (R) strategy is characterized by high fecundity, effective dispersal and short generation times.

We predict evolution for higher $b$, but not higher $c$ or lower $d$ (constrained to be large), amounts to focusing on fecundity. While there is no dispersal limitation in our model, propagules must be minimally effective at dispersal since our model only includes propagules which will enter territorial contests on unoccupied sites;

 Shorter generation times are thought to be a mechanism for increasing the chance that the reproductive cycle completes before disturbance-induced death

The evolution of the generation time itself is beyond the scope of our model, but it could

@@
